Presolving for\\ Mixed-Integer Semidefinite Optimization

### 2020-09-13 tex/MISDP-Presolving/MISDPpresolving.tex
some rework, added other Li and Xie inequalities

% title
\title{Presolving for\\ Mixed-Integer Semidefinite Optimization}
\author{Marc E. Pfetsch\thanks{Department of Mathematics, TU Darmstadt, Germany}}
\date{\today}

@@

The following inequalities are known from the literature, see
Gally~\cite{Gal19}, and can be added as linear inequalities w.r.t.\
\eqref{MISDP}:
\begin{itemize}
\item Any feasible $y \in \R^m$ satisfies $A(y) \succeq 0$, which implies
  that the diagonal entries of $A(y)$ are nonnegative, i.e., for all
  $i \in [n]$:
  \[
    \sum_{k = 1}^m (A^k)_{ii}\, y_k - (A^0)_{ii} \geq 0.
  \]
\item If for some $i$, $j \in [n]$, $A^0_{ij} \neq 0$, and $A^0_{ii} = 0$,
  $A^k_{ij} = 0$ for all $k \in [m]$, $A^k_{ii} = 0$ for all continuous
  variables and $\ell_i \geq 0$ for all integer variables,
  the following inequality is valid:
  \[
    \sum_{\substack{k \in \mathcal{I}:\\ A^k_{ii} > 0}} y_k \geq 1.
  \]
\item For all $X \succeq 0$, we have
  \[
    X_{ii} + X_{jj} - 2\, X_{ij} \geq 0,\qquad
    X_{ii} + X_{jj} + 2\, X_{ij} \geq 0.
  \]
  This follows by using a 2 by 2 minor and multiplying from left
  and right by the all-ones vector and $[1,-1]$, respectively. It also
  follows from~\eqref{eq:TwoByTwoMinorInequality}.
\item The previous inequalities translated to the matrix pencil are:
  \begin{align*}
    & \sum_{k=1}^m A^k_{ii}\, y_k - A^0_{ii} + \sum_{k=1}^m A^k_{jj}\, y_k - A^0_{jj} - 2 \Big(\sum_{k=1}^m A^k_{ij}\, y_k - A^0_{ij}\Big) \geq 0\\
    \quad\Leftrightarrow\quad&
    \sum_{k=1}^m \Big(A ^k_{ii} + A^k_{jj} - 2\, A^k_{ij}\Big)\, y_k \geq A^0_{ii} + A^0_{jj} - 2 A^0_{ij}.
  \end{align*}
\end{itemize}

All the above methods are implemented and help for some instances.

@@

\begin{proof}
  Since $X$ is positive semidefinite, we have
  \[
    X_{ii} X_{jj} - X_{ij}^2 \geq 0
    \quad\Leftrightarrow\quad
    X_{ij}^2 \leq X_{ii} X_{jj}.
  \]
  Moreover, since the diagonal terms of $X$ are nonnegative, we have
  $U_{ii} \geq L_{ii} \geq 0$ for all $i \in [n]$. Upper bounding via
  McCormick inequalities yields:
  \begin{align*}
    X_{ii} X_{jj} & \leq \min \{L_{jj}\, X_{ii} + U_{ii}\, X_{jj} - U_{ii}\, L_{jj},\;
                    U_{jj}\, X_{ii} + L_{ii}\, X_{jj} - L_{ii}\, U_{jj}\}  \\
    & \leq U_{ii}\, U_{jj}.
  \end{align*}
  Taking the square root gives the claim.
\end{proof}

@@

For a linear constraint, coefficient tightening can be explained by the
following example: Consider two binary variables~$x$ and $y$ and the
constraint $x + 2\, y \geq 1$. Then the coefficient of $y$ can be tightened
to $1$ yielding $x + y \geq 1$.

Assume that all matrices $A^k$, $k \in [m]$, are psd and $y_k$ is binary
for $k \in [m]$. Then we can compute
\[
  \hat{\mu}_k = \min\, \{ \mu \geq 0 \suchthat \mu A^k - A^0 \succeq 0\}
\]
adn define $\mu_k \define \min\, \{\hat{\mu}_k, 1\}$. Note that $\mu_k = 1$ if
$\hat{\mu}_k = \infty$, i.e., if the above optimization problem is infeasible.

@@

\begin{proof}
  For some matrix $A \in \R^{n \times n}$ and invertible matrix
  $B \in \R^{n \times n}$ we have $A \succeq 0$ if and only if
  $B\T A B \succeq 0$ (see, e.g., Helmberg~\cite[Prop.~1.1.7]{Hel00}).
  Thus, $\mu\, A^k - A^0 \succeq 0$ if and only if
  \begin{align*}
    0 \preceq (V^{-1})^\top (\mu\, A^k - A^0) V^{-1} &= \mu\,
    (V^{-1})^\top A^k V^{-1} - (V^{-1})^\top A^0 V^{-1} \\
    &= \mu\, I - (V^{-1})^\top A^0 V^{-1}.
  \end{align*}
  The eigenvalues of $\mu\, I - B$ for some matrix $B$ are $\mu - \lambda$
  for eigenvalues $\lambda$ of $B$.\footnote{This follows, for example, by the
    general theorem that if $g(t) \in \R[t]$ is a scalar polynomial and
    $\lambda_1, \dots, \lambda_n$ are the eigenvalues of $B$, then the
    eigenvalues of $g(B)$ are $g(\lambda_1), \dots, g(\lambda_n)$, see
    Gantmacher~\cite[Ch.~IV, Thm.~3]{Gan59I}. We here use
    $g(t) = \mu - t = \mu\, t^0 - t^1$. (Alternatively, $(\mu I - B)v =
    \lambda v \Leftrightarrow Bv = (\mu - \lambda)v$, so $v$ is an
    eigenvector for eigenvalue $\mu - \lambda$.)}  thus, to make
  $\mu\, I - V^{-\top} A^0 V^{-1}$ psd, we need
  $\mu \geq \lambda_{\max}(V^{-\top} A^0 V^{-1})$.
\end{proof}

@@

@@

Indeed, SOCP is special case of SDP, because:
\[
  (x,t) \in L_{n+1} \quad\Leftrightarrow\quad
  \tilde{X} \define
  \begin{pmatrix}
    t & x\T\\
    x & I_n
  \end{pmatrix}
  \succeq 0.
\]
To see this, we need the following result adapted
from~\cite[Section~A.5.5]{BoyV09}. Let
\[
  D = \begin{pmatrix}
    A & B\T\\
    B & C
  \end{pmatrix}
\]
with a positive definite $C \succ 0$. Then $D \succeq 0$ if and only if
$S = A - B\T C^{-1} B \succeq 0$ and $A \succeq 0$\footnote{Note that the
  condition $A \succeq 0$ is missing in~\cite{BoyV09}.}. For $D = \tilde{X}$, we
get: $\tilde{X} \succeq 0$ if and only if $t \geq 0$, $t - x\T x \geq
0$. The latter is equivalent to $(x,t) \in L_{n+!}$.

@@

\paragraph{A Valid SOCP Constraint}

Li and Xie~\cite{LiX20} show that the following inequalitiy is valid for
sparse PCA:
\[
  \sum_{j \in [n]} X_{ij}^2 \leq X_{ii}\, z_i \quad\text{for all } i \in [n].
\]
This holds, since in this case $X = x x\T$ and $\sum_{j=1}^m x_j^2 =
1$. Thus,
\[
  \sum_{j \in [n]} X_{ij}^2 \leq \sum_{j \in [n]} x_i^2\, x_j^2 = x_i^2
  \leq z_i\, X_{ii}.
\]
Note that these inequalities can be reformulated by subtracting $X_{ii}^2$
from both sides to yield
\[
  \sum_{\substack{j \in [n]\\ j \neq i}} X_{ij}^2 \leq z_i\, (X_{ii} - X_{ii}^2).
\]

@@

>>>>

% title
\title{Presolving for\\ Mixed-Integer Semidefinite Optimization}
\author{Marc E. Pfetsch\thanks{Department of Mathematics, TU Darmstadt, Germany}}
\date{September 2020}

@@

The following inequalities are known from the literature, see
Mars~\cite{Mar13} and  Gally~\cite{Gal19}, and can be added as linear inequalities w.r.t.\
\eqref{MISDP}:
\begin{itemize}
\item Any feasible $y \in \R^m$ satisfies $A(y) \succeq 0$, which implies
  that the diagonal entries of $A(y)$ are nonnegative, i.e., for all
  $i \in [n]$:
  \[
    \sum_{k = 1}^m (A^k)_{ii}\, y_k - (A^0)_{ii} \geq 0.
  \]
\item If for some $i$, $j \in [n]$, $A^0_{ij} \neq 0$, and $A^0_{ii} = 0$,
  $A^k_{ij} = 0$ for all $k \in [m]$, $A^k_{ii} = 0$ for all continuous
  variables and $\ell_i \geq 0$ for all integer variables,
  the following inequality is valid:
  \[
    \sum_{\substack{k \in \mathcal{I}:\\ A^k_{ii} > 0}} y_k \geq 1.
  \]
  Note that the conditions imply that $A^0$ is not psd.
\item For all $X \succeq 0$, we have
  \[
    X_{ii} + X_{jj} - 2\, X_{ij} \geq 0,\qquad
    X_{ii} + X_{jj} + 2\, X_{ij} \geq 0.
  \]
  This follows by using a 2 by 2 minor and multiplying from left
  and right by the all-ones vector and $[1,-1]$, respectively. It also
  follows from~\eqref{eq:TwoByTwoMinorInequality}.
\item The previous inequalities translated to the matrix pencil are:
  \begin{align*}
    & \sum_{k=1}^m A^k_{ii}\, y_k - A^0_{ii} + \sum_{k=1}^m A^k_{jj}\, y_k - A^0_{jj} - 2 \Big(\sum_{k=1}^m A^k_{ij}\, y_k - A^0_{ij}\Big) \geq 0\\
    \quad\Leftrightarrow\quad&
    \sum_{k=1}^m \Big(A ^k_{ii} + A^k_{jj} - 2\, A^k_{ij}\Big)\, y_k \geq A^0_{ii} + A^0_{jj} - 2 A^0_{ij}.
  \end{align*}
\end{itemize}

All the above methods are implemented and seem to help for some instances.

@@

\begin{proof}
  Since $X$ is positive semidefinite, we have
  \[
    X_{ii} X_{jj} - X_{ij}^2 \geq 0
    \quad\Leftrightarrow\quad
    X_{ij}^2 \leq X_{ii} X_{jj}.
  \]
  Moreover, since the diagonal terms of $X$ are nonnegative, we have
  $U_{ii} \geq L_{ii} \geq 0$ for all $i \in [n]$. Thus,

  Upper bounding via
  McCormick inequalities yields:
  \begin{align*}
    X_{ii} X_{jj} & \leq \min \{L_{jj}\, X_{ii} + U_{ii}\, X_{jj} - U_{ii}\, L_{jj},\;
                    U_{jj}\, X_{ii} + L_{ii}\, X_{jj} - L_{ii}\, U_{jj}\}  \\
    & \leq \min \{L_{jj}\, U_{ii} + U_{ii}\, U_{jj} - U_{ii}\, L_{jj},\;
                    U_{jj}\, U_{ii} + L_{ii}\, U_{jj} - L_{ii}\, U_{jj}\}  \\
    & = \min \{U_{ii}\, U_{jj},\;  U_{jj}\, U_{ii}\} = U_{ii}\, U_{jj}.
  \end{align*}
  Taking the square root gives the claim.
\end{proof}

@@

For a linear constraint, coefficient tightening can be explained by the
following example: Consider two binary variables~$x$ and $y$ and the
constraint $x + 2\, y \geq 1$. Then the coefficient of $y$ can be tightened
to $1$, yielding $x + y \geq 1$.

Assume that all matrices $A^k$, $k \in [m]$, are psd and $y_k$ is binary
for $k \in [m]$. Then we can compute
\begin{equation}\label{eq:ComputeHatMu}
  \hat{\mu}_k = \min\, \{ \mu \geq 0 \suchthat \mu A^k - A^0 \succeq 0\}
\end{equation}
and define $\mu_k \define \min\, \{\hat{\mu}_k, 1\}$. Note that $\mu_k = 1$ if
$\hat{\mu}_k = \infty$, i.e., if the above optimization problem is infeasible.

@@

\begin{proof}
  For some matrix $A \in \R^{n \times n}$ and invertible matrix
  $B \in \R^{n \times n}$ we have $A \succeq 0$ if and only if
  $B\T A B \succeq 0$ (see, e.g., Helmberg~\cite[Prop.~1.1.7]{Hel00}).
  Thus, $\mu\, A^k - A^0 \succeq 0$ if and only if
  \begin{align*}
    0 \preceq (V^{-1})^\top (\mu\, A^k - A^0) V^{-1} &= \mu\,
    (V^{-1})^\top A^k V^{-1} - (V^{-1})^\top A^0 V^{-1} \\
    &= \mu\, I - (V^{-1})^\top A^0 V^{-1}.
  \end{align*}
  The eigenvalues of $\mu\, I - B$ for some matrix $B$ are $\mu - \lambda$
  for eigenvalues $\lambda$ of $B$.\footnote{This follows, for example, by the
    general theorem that if $g(t) \in \R[t]$ is a scalar polynomial and
    $\lambda_1, \dots, \lambda_n$ are the eigenvalues of $B$, then the
    eigenvalues of $g(B)$ are $g(\lambda_1), \dots, g(\lambda_n)$, see
    Gantmacher~\cite[Ch.~IV, Thm.~3]{Gan59I}. We here use
    $g(t) = \mu - t = \mu\, t^0 - t^1$. (Alternatively, $(\mu I - B)v =
    \lambda v \Leftrightarrow Bv = (\mu - \lambda)v$, so $v$ is an
    eigenvector for the eigenvalue $\mu - \lambda$.)}  Thus, to make
  $\mu\, I - V^{-\top} A^0 V^{-1}$ psd, we need
  $\mu \geq \lambda_{\max}(V^{-\top} A^0 V^{-1})$.
\end{proof}

@@

Note that~\eqref{eq:ComputeHatMu} might be infeasible, e.g., if $A^0$ is
not negative definite and $A^k$ is psd with minimal eigenvalue 0. An
extreme example is if $A^k = 0$, but $A^0 \neq 0$. Thus, the implementation
of these methods have to be safe-guarded against infeasibility.

\begin{question}
  Is there an easy way to determine whether~\eqref{eq:ComputeHatMu} is
  infeasible, e.g., by computing eigenvalues once?  \footnote{For example
    is the following true: If $\mu A^k - A^0$ is not psd for
    $\mu \geq \lambda_{\max}(A^0)/\lambda_{\max}(A^k)$ then~\eqref{eq:ComputeHatMu} is
    infeasible? This would require that $\mu A^k - A^0 \succeq 0$ for all
    $\mu \geq \lambda_{\max}(A^0) / \lambda_{\max}(A^k)$, e.g., if $A^0$ is
    psd and if there exists any $\mu'$ such that $\mu' A^k - A^0 \succeq 0$.}
\end{question}

@@

Indeed, SOCP is special case of SDP, because:
\[
  (x,t) \in L_{n+1} \quad\Leftrightarrow\quad
  \tilde{X} \define
  \begin{pmatrix}
    t & x\T\\
    x & I_n
  \end{pmatrix}
  \succeq 0.
\]
To see this, we need the following result adapted
from~\cite[Section~A.5.5]{BoyV09}. Let
\[
  D = \begin{pmatrix}
    A & B\T\\
    B & C
  \end{pmatrix}
\]
with a positive definite $C \succ 0$. Then $D \succeq 0$ if and only if
$S = A - B\T C^{-1} B \succeq 0$ and $A \succeq 0$\footnote{Note that the
  condition $A \succeq 0$ is missing in~\cite{BoyV09}.}. For $D = \tilde{X}$, we
get: $\tilde{X} \succeq 0$ if and only if $t \geq 0$, $t - x\T x \geq
0$. The latter is equivalent to $(x,t) \in L_{n+1}$.

@@

\paragraph{Two Valid SOCP Constraint}

Li and Xie~\cite{LiX20} show that the following inequalitiy is valid for
sparse PCA:
\[
  \sum_{j \in [n]} X_{ij}^2 \leq X_{ii}\, z_i \quad\text{for all } i \in [n].
\]
This holds, since in this case on can assume that $X = x x\T$ and $\sum_{j=1}^m x_j^2 =
1$.\footnote{Note that this might cut off optimal solutions that do not
  have rank 1.} Thus,
\[
  \sum_{j \in [n]} X_{ij}^2 \leq \sum_{j \in [n]} x_i^2\, x_j^2 = x_i^2
  \leq z_i\, X_{ii}.
\]
Note that these inequalities can be reformulated by subtracting $X_{ii}^2$
from both sides to yield
\[
  \sum_{\substack{j \in [n]\\ j \neq i}} X_{ij}^2 \leq z_i\, (X_{ii} - X_{ii}^2).
\]

@@

A second type of inequality of Li and Xie~\cite{LiX20} is
\[
  \bigg(\sum_{j \in [n]} \abs{X_{ij}}\bigg)^2 \leq k\, X_{ii}\, z_i\quad
  \forall\, i \in [n].
\]
These inequalities seem to be nontrivial
\footnote{Note that the inequality $\sum_{j \in [n]} \abs{X_{ij}} \leq k$ is trivial,
since $\abs{X_{ij}} \leq z_j$ and thus
\[
  \sum_{j \in [n]} \abs{X_{ij}} \leq \sum_{j \in [n]} z_j \leq k.
\]
Similarly, $\sum_{j \in [n]} \abs{X_{ij}} \leq z_i$ holds, but it seems
that $\sum_{j \in [n]} \abs{X_{ij}} \leq X_{ii}\, z_i$ does not hold in general.}
and follow, because
\[
  \sum_{j \in [n]} \abs{X_{ij}} = \sum_{j \in [n]} \abs{x_i} \abs{x_j} =
  \abs{x_i} \sum_{j \in [n]} \abs{x_j} \leq \abs{x_i} \sqrt{k},
\]
where we used that $\norm{x}_1 \leq \sqrt{k}$ if $\norm{x}_2 = 1$ and
$\norm{x}_0 \leq k$. Moreover, $\abs{x_i} = \sqrt{X_{ii}} = \sqrt{X_{ii}
  z_i}$, because if $z_i = 0$ then $X_{ii} = 0$.

These inequalities are SOC representable, but one needs to express the
absolute values in $\sum_{j \in [n]} \abs{X_{ij}}$, e.g., by variable
splitting.
